Computing Entropies With Nested Sampling

### 2017-08-11 paper/response.tex

I'm glad you liked it! If you found this interesting, check out the foundations
work of Kevin Knuth, which I find fascinating.

{\quote
The author also writes:
"The mutual information is another way of describing the relevance of the data to the parameters. its definition, and relation to other quantities"
As far as I am aware, this is a novel way of thinking about mutual information!  Very nice!
In addition, I did not know that it could be written as the expected value of KL over datasets, which is fascinating.}

Thanks! In most information theory textbooks, these things are written in
terms of abstract `random variables' $X$ and $Y$. One of my subsidiary
goals in this paper was to discuss this stuff in terms and notation more
familiar to Bayesians.

{\quote
4. It would be useful to follow (13) with the calculation written in your notation rather than Skilling's.}

I decided to do this the other way around, first writing it in my notation and
then switching to Skilling's, which is more convenient when the computational
algorithm is being discussed.

{\quote
5. After (13): Why does the implied distribution of $L$-values tend to be heavy-tailed?}

Imagine doing simple Monte Carlo to get $Z$, by drawing from the prior and
collecting likelihood values, which you'll want to average. A histogram of
those likelihood values will look ridiculous, and you might have to wait
a million years until you get enough points such that your average becomes
accurate, because it will be dominated by huge likelihoods which are seen
rarely. Most NS people talk about this in terms of the parameter space, but
it also has an interpretation in terms of a `probability distribution for
$L$' which is implied by $\pi$.

{\quote
6. Just before (14) and in (15): I do not understand the notation regarding the implicit likelihood boundary involving \\mathbb\{1\}}

I've added a bit of explanation of the indicator function notation.

### 2017-08-11 paper/ms.tex
More editing

Nested Sampling (NS), introduced by
\citet{skilling2006nested}, is an algorithm whose aim is
to calculate the evidence
\begin{align}
Z &= \int \pi(\x) L(\x) \, d\x
\end{align}
where $\x$ is the unknown parameter(s), $\pi$ is the prior distribution,
and $L$ is the likelihood function. The original NS algorithm, and variations
within the same family \citep{feroz2009multinest, dns, handley2015polychord},
have become popular tools in Bayesian data analysis
\citep{knuth2015bayesian, pullen2014bayesian, exoplanet} and have also
found use in statistical physics \citep{partay2010efficient,
baldock2016determining, martiniani2014superposition}, which was another of
its aims.
{\revision
NS also estimates the Kullback-Leibler divergence from the prior
$\pi(\x)$ to the posterior $\pi(x)L(x)/Z$, which Skilling calls the
information. This is a measure of how much was learned about $\x$
from the specific dataset, and is also useful for defining a termination
criterion for NS.}

While $Z$ is a simple expectation with
respect to $\pi$, the
implied distribution of $L$-values tends to be very heavy-tailed, which is
why simple Monte Carlo does not work. Equivalently, the integral is dominated
by very small regions where $L(\x)$ is high.

To overcome this, NS evolves a population of $N$ particles in the parameter
space towards higher likelihood regions.
The particles are initially initialised from the prior $\pi(\x)$, and the
particle with the lowest likelihood, $L^*_1$, is found and its details are
recorded as output. This worst particle is then discarded
and replaced by a new particle
drawn from the prior $\pi$ but subject to the constraint that its likelihood
must be above $L^*_1$ (this is usually achieved by cloning a surviving particle
and evolving it with MCMC to explore
$\pi(\x)\mathds{1}\left(L(\x) > L^*_1\right)$).
This process is repeated, resulting in an
increasing sequence of likelihoods
\begin{align}
L^*_1, L^*_2, L^*_3,... \label{eqn:likelihood_sequence}
\end{align}
Defining $X(\ell)$ as the amount of prior mass with likelihood greater than
some threshold $\ell$,
\begin{align}
X(\ell) &= \int \pi(\x) \mathds{1}\left(L(\x) > \ell\right) \, d\x,
\end{align}
each discarded point in the sequence can be assigned an $X$-value,
which transforms $Z$ to a one-dimensional integral.
\citet{skilling2006nested}'s key insight was that the
$X$-values of the discarded points are approximately known from the nature of the
algorithm. Specifically, each iteration reduces the prior mass by approximately
a factor $e^{-1/N}$. More accurately, the conditional distribution of
$X_{i}$ given $X_{i-1}$ (and the definition of the algorithm)
is obtainable by letting
$t_i \sim \textnormal{Beta}(N,1)$ and setting $X_{i+1} := t_iX_{i-1}$ (where
$X_1 = t_1$).
The $t_i$ quantities are the proportion of the remaining prior mass
that is {\em retained} at each iteration, after imposing the constraint of the
latest $L^*$ value.

>>>>

Nested Sampling (NS), introduced by
\citet{skilling2006nested}, is an algorithm whose aim is
to calculate the evidence {\revision
\begin{align}
p(\x) &= \int p(\x) p(\d | \x) \, d\x,
\end{align}
or, in the simplified notation common when discussing computational matters,
\begin{align}
Z &= \int \pi(\x) L(\x) \, d\x,
\end{align}}
\noindent where $\x$ is the unknown parameter(s), $\pi$ is the prior distribution,
and $L$ is the likelihood function. The original NS algorithm, and variations
within the same family \citep{feroz2009multinest, dns, handley2015polychord},
have become popular tools in Bayesian data analysis
\citep{knuth2015bayesian, pullen2014bayesian, exoplanet} and have also
found use in statistical physics \citep{partay2010efficient,
baldock2016determining, martiniani2014superposition}, which was another of
its aims.
{\revision
NS also estimates the Kullback-Leibler divergence from the prior
$\pi(\x)$ to the posterior $\pi(x)L(x)/Z$, which Skilling calls the
information. This is a measure of how much was learned about $\x$
from the specific dataset, and is also useful for defining a termination
criterion for NS.}

While $Z$ is a simple expectation with
respect to $\pi$, the
{\revision distribution of $L$-values implied by $\pi$}
tends to be very heavy-tailed, which is
why simple Monte Carlo does not work. Equivalently, the integral 
{\revision over $\x$} is dominated by very small regions where $L(\x)$ is high.

To overcome this, NS evolves a population of $N$ particles in the parameter
space towards higher likelihood regions.
The particles are initially initialised from the prior $\pi(\x)$, and the
particle with the lowest likelihood, $L^*_1$, is found and its details are
recorded as output. This worst particle is then discarded
and replaced by a new particle
drawn from the prior $\pi$ but subject to the constraint that its likelihood
must be above $L^*_1$. {\revision This is usually achieved by cloning a surviving particle
and evolving it with MCMC to explore the prior distribution but with
a constraint on likelihood value, $L(\x) > L^*_1$. This `constrained prior'
distribution has probability density proportional to
\begin{align}
\pi(\x)\mathds{1}\left(L(\x) > L^*_1\right)
\end{align}
where $\mathds{1}()$ is an indicator function which returns one if the argument
is true and zero if it is false.}

This process is repeated, resulting in an
increasing sequence of likelihoods
\begin{align}
L^*_1, L^*_2, L^*_3,... \label{eqn:likelihood_sequence}
\end{align}
Defining $X(\ell)$ as the amount of prior mass with likelihood greater than
some threshold $\ell$,
\begin{align}
X(\ell) &= \int \pi(\x) \mathds{1}\left(L(\x) > \ell\right) \, d\x,
\end{align}
each discarded point in the sequence can be assigned an $X$-value,
which transforms $Z$ to a one-dimensional integral.
\citet{skilling2006nested}'s key insight was that the
$X$-values of the discarded points are approximately known from the nature of the
algorithm. Specifically, each iteration reduces the prior mass by approximately
a factor $e^{-1/N}$. More accurately, the conditional distribution of
$X_{i}$ given $X_{i-1}$ (and the definition of the algorithm)
is obtainable by letting
$t_i \sim \textnormal{Beta}(N,1)$ and setting $X_{i+1} := t_iX_{i-1}$ (where
$X_1 = t_1$).
The $t_i$ quantities are the proportion of the remaining prior mass
that is {\em retained} at each iteration, after imposing the constraint of the
latest $L^*$ value.
